---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

commanism <- function(.) {
    scales::label_comma(accuracy = 1)(.)
}

set.seed(42)
```

# hercacstables

This package provides an R interface to the United States Census Bureau's data
API.
The primary focus, as per its name, is pulling information from the
[detailed tables of the American Community Survey](https://www.census.gov/programs-surveys/acs/data/data-tables.html).

<!-- badges: start -->
<!-- badges: end -->

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("higherX4Racine/hercacstables")
```

## Fetching data from the API

The main way that one uses `hercacstables` to interact with the Census API is
the `fetch_data()` function.

## A vanilla call to `fetch_data()`

Here is a modestly complicated use of the function without any setup or
post-processing.

```{r}
#| label: fetch-pops-and-households
#| cache: true

POPS_AND_HOUSEHOLDS <- hercacstables::fetch_data(
    # the API works one year at a time
    year = hercacstables::most_recent_vintage("acs", "acs1"),
    
    # the API can only query one data source at a time
    survey_type = "acs",           # the American Community Survey
    table_or_survey_code = "acs1", # the 1-year dataset of the ACS
    
    # the API fetches values for one or more instances of a specific geography
    for_geo = "state",             # fetch values for entire states
    for_items = c(
        "11",                      # the District of Columbia
        "72"                       # Puerto Rico
    ),
    
    # the codes for the specific data that we are fetching
    variables = c(
        "NAME",                    # the geographic area's name
        "B01001_001E",             # the total number of people
        "B11002_002E",             # people in family households
        "B11002_012E"              # people in nonfamily households
    )
)
```

```{r}
#| label: show-pops-and-households
#| echo: false
POPS_AND_HOUSEHOLDS |>
    dplyr::mutate(
        Value = commanism(.data$Value)
    ) |>
    knitr::kable(
        align = "lrlrrr"
    )
```

## Identifying which items to fetch

People working with the Census often know the topic that they are interested in,
but not the specific variables that provide information about that topic.
The `hercacstables` package has utilities that help you search for Census
variables if you don't already know which ones you want to use.

### Search for groups with keywords

A good first step is to search for tables that could be relevant with the
`search_in_glossary()` function and the built-in
[`METADATA_FOR_ACS_GROUPS`](reference/METADATA_FOR_ACS_GROUPS.html).

The following example searches using three columns.
The column "Group" needs to end in a digit.
That will limit the results to tables that report numbers for all races.
The column "Universe" needs to include the string "population".
That will limit the results to tables that report counts of people.
The column "Description" needs to both start with the string "school" and
include the string "enroll".
This will limit the results to tables about enrollment in school.

```{r}
#| label: search-in-glossary
EDUCATION_TABLES <- hercacstables::search_in_glossary(
    hercacstables::METADATA_FOR_ACS_GROUPS,
    Group = "\\d$",
    Universe = "population",
    Description = c("^school", "enroll")
)
```

```{r}
#| label: show-glossary_search-results
#| echo: false
knitr::kable(EDUCATION_TABLES)
```

### Unpack variables for a group

Once you know the group that you are interested in, you will want to see what
information is captured in each of its rows.
The `unpack_group_details()` function searches the built-in
[`METADATA_FOR_ACS_VARIABLES`](reference/METADATA_FOR_ACS_VARIABLES)
for all of the variables of one group.
It then expands the Census's description of each variable into columns.
Users will need to identify the concepts that are captured by each column.

The following example unpacks the variables in the [B14001](https://api.census.gov/data/2023/acs/acs1/B14001.html) table.

```{r}
#| label: unpack-B14001
UNPACKED_B14001 <- hercacstables::unpack_group_details("B14001")
```

```{r}
#| label: show-B14001
#| echo: false
UNPACKED_B14001 |>
    dplyr::filter(.data$Dataset == "ACS1") |>
    dplyr::select(!"Dataset") |>
    knitr::kable()
```

## Shortcuts

The package also includes functions for common special cases. 
They are wrappers for `fetch_data()` with some arguments hard-coded.

### Decennial populations by race and ethnicity

One example is pulling trends of racial/ethnic populations from the decennial
census for some specific geographical units.

```{r}
#| label: fetch-decennial-pops-by-race
#| cache: true
POPS_BY_RACE <-
    hercacstables::fetch_decennial_pops_by_race(
        for_geo = "state", # one cannot fetch the whole nation from 2000 or 2010
        for_items = "*"    # so we pull the data for every state
    ) |>
    dplyr::count( # then compute the nationwide populations for each vintage
        .data$`Race/Ethnicity`,
        .data$Vintage,
        wt = .data$Population,
        name = "Population"
    ) |>
    tidyr::pivot_wider( # and reshape the table for display
        names_from = "Vintage",
        values_from = "Population"
    )
```

```{r}
#| label: show-decennial-pops-by-race
#| echo: false
POPS_BY_RACE |>
    dplyr::mutate(
        dplyr::across(tidyselect::ends_with("0"),
                      commanism)
    ) |>
    knitr::kable(
        align = "lrrr"
    )
```
