---
title: "Race/Ethnicity Categories in ACS Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Race/Ethnicity Categories in ACS Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
library(hercacstables)
```

## Preface

The American Community Survey API returns data in a very terse format.
Each response contains four parts: geographic information, a table code, a row
number, and a value.
The geographic information varies by the level of detail one asks for.
The value is a number that may be a population size, number of households, an
income in dollars, a percentage, or several other quantities.
The table code and row number are what determine the exact meaning of the value.

Generally speaking, all of the values in a table will be reporting the same kind
of information.
Again, the information might be counts of people or households, incomes or costs
in dollars, hours commuting or working, or a percentage of just about anything.
The row numbers then let you know what group of people the information is for.
The first row always has a value for the entire population of a geographic area.
Subsequent rows may have information about a very specific subgroup, or may
contain a summary value for a combination of subgroups.
For example, in a sex-and-age table, you might find the value for all men and
boys (males) in row 2, then the value for boys under 5 in row 3.
In short, a lot of information is packed into the two fields of table name and
row number.

While packing information is efficient for storing and transmitting data, it
means that users of Census information must unpack things in order to be clear.
The `hercacstables` package has many tools and metadata tables that should make
it easier and more convenient to unpack Census data.

## Racial and ethnic categories in the American Community Survey

This vignette describes some features of `hercacstables` that help with
the common and repetitive chores of unpacking racial and ethnic data from
Census API responses.
The Census has several ways that it reports by racial and ethnic identity.
They do not always all agree or line up.
That makes sense because the concepts of race and ethnicity are slippery,
changing across place and time.

### Broad strokes by the Census

The first racial/ethnic system that we will discuss involves ten categories.
There are seven categories of race, one of ethnicity, one that combines both,
and an "All" category:
"Total",
"White alone",
"Black or African American alone",
"American Indian and Alaska Native alone",
"Asian alone",
"Native Hawaiian and Other Pacific Islander alone",
"Some Other Race alone",
"Two or More Races",
"White alone, not Hispanic or Latino",
and "Hispanic or Latino".
This is a very coarse and unsophisticated way of characterizing identities.
Nevertheless, it occurs throughout the American Community Survey.

### Subtable categories of race and ethnicity

```{r}
#| label: find-subtables-with-ethnicity
#| echo: false

broad_subtables <- METADATA_ACS5 |>
    purrr::pluck("groups") |>
    dplyr::filter(stringr::str_detect(name, "I$")) |>
    dplyr::select("name", "description") |>
    dplyr::mutate(
        name = stringr::str_remove(.data$name, "I$"),
        description = .data$description |>
            stringr::str_remove("\\(HISPANIC OR LATINO[^)]*\\)") |>
            stringr::str_squish()
    )
```

There are `r nrow(broad_subtables)`
[separate sets of tables](#subtables-of-race-and-ethnicity) that subdivide their
information according to this ten-category scheme.
Each identity is designated by a suffix at the end of a table's name.
In other words, all of the values from a table with one of these suffixes in its
name will pertain to people of just one racial/ethnic category.

Since the case is so common, `hercacstables` describes it with the metadata
table `RACE_ETHNICITY_SUBTABLE_METADATA`.
It maps from the suffix in a table's name to the racial/ethnic category the
table describes.

```{r}
#| label: show-race-ethnicity-subtable-metadata
#| echo: false
RACE_ETHNICITY_SUBTABLE_METADATA |>
    dplyr::select(
        "Census Race",
        "Suffix"
    ) |>
    knitr::kable()
```

### Tables with "RACE" in their description

```{r}
#| label: find-race-in-table-descriptions
#| echo: false
tables_that_name_race <- METADATA_ACS5$groups |>
    dplyr::filter(
        stringr::str_detect(.data$name, "\\d$"),
        stringr::str_detect(.data$description, "RACE")
    ) |>
    dplyr::select(
        "name", "description"
    ) |>
    dplyr::left_join(
        dplyr::count(METADATA_ACS5$variables,
                     .data$group,
                     name = "rows"),
        by = c(name = "group")
    )
```

There are also `r nrow(tables_that_name_race)` tables that have the word "RACE"
in their description.


```{r}
#| label: show-the-tables-that-name-race
#| echo: false
knitr::kable(tables_that_name_race)
```

They seem to fall into four categories.
There are two tables with 10 rows each, `B02001` and `B25006`.
These are probably convenience tables that pull data from other race-specific
subtables.
Six tables, `B02008` through `B020013`, have only one row each.
These seem to report inclusive counts, combining people who claim a
racial identity either as their sole identity or in combination with another.
One table, `B03002`, has 21 rows.
This table appears to report detailed information about Hispanic ethnicity and
specific racial identities.
Tables `B98013` and `B99021` have descriptions that suggest that they report
methodological details.
Each of these categories of table deserves a little more discussion.

#### Convenience tables

Tables `B02001` and `B25006` deal with counts of individuals and households,
respectively.
They seem to be redundant, showing the first rows of the tables in the `B01001`
and `B11001`, respectively.

We can use `hercacstables` to check that.

##### Map meaning to Census Variables

The first step is to make a table that connects the Census's opaque variable
names to the real-world meanings that we are interested in.
For example, tables `B01001*` and `B02001` deal with counts of people,
while tables `B11001*` and `B25006` deal with counts of households.
Similarly, the racial identity being counted is defined either by the subtable
suffix or the row number.
Finally, we are interested in whether the data come from, a subtable or a
convenience table.
We can lay that all out in a way that "maps" from the Census variable to the
real-world meaning.

```{r}
#| label: check-variables-for-convenience-tables

convenience_check_variables <- tibble::tribble(
    ~ Table,  ~ Suffix,   ~ Index, ~ Population, ~ Race,
    "B01001", "A",              1, "People",     "White",
    "B01001", "B",              1, "People",     "Black",
    "B02001", "",               2, "People",     "White",
    "B02001", "",               3, "People",     "Black",
    "B11001", "A",              1, "Households", "White",
    "B11001", "B",              1, "Households", "Black",
    "B25006", "",               2, "Households", "White",
    "B25006", "",               3, "Households", "Black"
) |>
    dplyr::mutate(
        Group = paste0(.data$Table, .data$Suffix),
        Variable = build_api_variable(group_code = .data$Table,
                                      race_code = .data$Suffix,
                                      item_number = .data$Index),
        `Table Type` = dplyr::if_else(nchar(.data$Suffix) > 0,
                                      "Subtable",
                                      "Convenience")
    )

knitr::kable(convenience_check_variables)
```

##### Fetch the raw data

With our variables defined, we can fetch the data from the Census API.
Notice that, since this is an API call, the code block is set to cache its
results.
API calls are much slower than local functions, so it is usually a good idea to
isolate them and run them as few times as possible.

```{r, cache=TRUE}
#| label: check-values-for-convenience-tables

convenience_check_values_raw <- fetch_data(
    convenience_check_variables$Variable,
    year = 2022,                  # the most recent one available at the time of writing
    for_geo = "us",               # the entire nation
    for_items = "*",              # all nation-level geographies
    survey_type = "acs",          # as opposed to, e.g. "dec," for Decennial survey data
    table_or_survey_code = "acs5" # the specific survey is the 5-year ACS.
)
```

##### Examine values in context

Now that we have the raw data, we can check to see if the values from the
convenience tables do, in fact, match up with the values from the subtables.

```{r}
#| label: wrangle-convenience-check-data

convenience_check_values <- convenience_check_values_raw |>
    dplyr::inner_join(
        convenience_check_variables,
        by = c("Group", "Index")
    ) |>
    dplyr::select(
        "Population",
        "Race",
        "Table Type",
        "Value"
    ) |>
    tidyr::pivot_wider(
        names_from = "Table Type",
        values_from = "Value"
    ) |>
    dplyr::mutate(
        Identical = dplyr::if_else(.data$Subtable == .data$Convenience,
                                   "Yes",
                                   "No")
    )

convenience_check_values |>
    dplyr::mutate(
        dplyr::across(tidyselect::all_of(c("Subtable", "Convenience")),
                      scales::label_comma(accuracy = 1))
    ) |>
    knitr::kable(
        align = "lrrl"
    )
```

We don't have to include any of the confusing "Group," "Index," or "Variable"
columns in our final result.

#### Inclusive identity counts

Six others, tables "B02008" through "B02013", show the numbers of people who
identified with specific races and ethnicities.
The totals from these tables will be larger than the US population because
someone who identified with more than one category will be counted in each
corresponding table.

These tables correspond to six of the ten broad categories of race/ethnicity, so
they are actually already in `RACE_ETHNICITY_SUBTABLE_METADATA`.
I just hid them before because it would have been confusing.

```{r}
#| label: show-inclusive-columns-too
#| echo: false
RACE_ETHNICITY_SUBTABLE_METADATA |>
    dplyr::select(
        "Census Race",
        "Suffix",
        "Inclusive Table"
    ) |>
    knitr::kable()
```

The population reported in the inclusive tables should be as large, or larger,
than the population reported in the exclusive tables.
We can use `hercacstables` to check this, too.

##### Census variables for inclusive and exclusive counts

```{r}
#| label: inclusive-exclusive-population-variables
incl_excl_pop_variables <- RACE_ETHNICITY_SUBTABLE_METADATA |>
    dplyr::filter(
        nchar(.data$`Inclusive Table`) > 0
    ) |>
    dplyr::mutate(
        `Exclusive Table` = paste0("B01001", .data$Suffix)
    ) |>
    dplyr::select(
        "Census Race",
        Inclusive = "Inclusive Table",
        Exclusive = "Exclusive Table"
    ) |>
    tidyr::pivot_longer(
        cols = tidyselect::ends_with("clusive"),
        names_to = "Table",
        values_to = "Group"
    ) |>
    dplyr::mutate(
        Variable = build_api_variable(.data$Group, 1)
    )

knitr::kable(incl_excl_pop_variables)
```

##### Raw inclusive and exclusive counts

```{r, cache=TRUE}
#| label: fetch-inclusive-exclusive-populations
incl_excl_pop_values_raw <- fetch_data(
    variables = incl_excl_pop_variables$Variable,
    year = 2022,
    for_geo = "us",
    for_items = "*",
    survey_type = "acs",
    table_or_survey_code = "acs5"
)
```

##### Comparing inclusive and exclusive counts

```{r}
#| label: wrangle-inclusive-exclusive-populations
incl_excl_pop_values <- incl_excl_pop_values_raw |>
    dplyr::inner_join(
        incl_excl_pop_variables,
        by = c("Group")
    ) |>
    dplyr::select(
        "Census Race",
        "Table",
        "Value"
    ) |>
    tidyr::pivot_wider(
        names_from = "Table",
        values_from = "Value"
    ) |>
    dplyr::mutate(
        Difference = .data$Inclusive - .data$Exclusive,
        `Percent Multiracial` = .data$Difference / .data$Inclusive
    ) |>
    dplyr::arrange(
        dplyr::desc(.data$Inclusive)
    )

incl_excl_pop_values |>
    dplyr::mutate(
        dplyr::across(tidyselect::all_of(c("Inclusive",
                                           "Exclusive",
                                           "Difference")),
                      scales::label_comma(accuracy = 1)),
        `Percent Multiracial` = scales::label_percent(accuracy = 1)(
            .data$`Percent Multiracial`
        )
    ) |>
    knitr::kable(
        align = "lrrrr"
    )
```

#### Hispanic ethnicity and broad racial identity

Table `B03002` contains counts of people by Hispanic ethnicity for each of the
ten broad racial identities.
As always, the first row is the total population size.
There are then two groups of rows.
Rows 2-11 count people who are not Hispanic.
Rows 12-21 count people who identify as Hispanic.
Rows 2 and 12 are the total populations of non-Hispanic and Hispanic people.
Rows 10, 11, 20, and 21 contain subgroupings that distinguish between people who 
identify as biracial and those who identify as multiracial.

That means that we can take rows one through eight of
`RACE_ETHNICITY_SUBTABLE_METADATA` and map them onto rows in `B03002`.
In fact, this is so useful that it is also included in
`RACE_ETHNICITY_SUBTABLE_METADATA`.

```{r}
#| label: hispanic-and-broad-race-variables
#| echo: false

RACE_ETHNICITY_SUBTABLE_METADATA |>
    dplyr::filter(
        nchar(.data$`non-Hispanic`) > 0
    ) |>
    dplyr::select(
        "Census Race",
        "non-Hispanic",
        "Hispanic"
    ) |>
    knitr::kable()
```

Let's use these to look at nationwide trends across racial identities in their
percentages of Hispanic ethnicity.

First, we will define our metadata table.

```{r}
#| label: define-hispanic-and-broad-race-variables

hispanic_and_broad_race_variables <- RACE_ETHNICITY_SUBTABLE_METADATA |>
    dplyr::filter(
        nchar(.data$`non-Hispanic`) > 0
    ) |>
    dplyr::select(
        "Census Race",
        "non-Hispanic",
        "Hispanic"
    ) |>
    tidyr::pivot_longer(
        cols = tidyselect::ends_with("Hispanic"),
        names_to = "Ethnicity",
        values_to = "Variable"
    ) |>
    tidyr::separate_wider_position(
        cols = "Variable",
        widths = c(Group = 6, 1,
                   Index = 3,
                   1),
        cols_remove = FALSE
    ) |>
    dplyr::mutate(
        Index = as.integer(.data$Index)
    )

knitr::kable(hispanic_and_broad_race_variables)
```

Next, we pull 13 years of ACS data.

```{r, cache=TRUE}
#| label: fetch-hispanic-and-broad-race-data

hispanic_and_broad_race_raw <- c(2009:2019, 2021:2022) |>
    purrr::map(
        ~ fetch_data(
            hispanic_and_broad_race_variables$Variable,
            year = .,
            for_geo = "us",
            for_items = "*",
            survey_type = "acs",
            table_or_survey_code = "acs1"
        )
    )
```

Then, we put the data into a nice, tidy format.

```{r}
#| label: tidy-hispanic-and-broad-race-data
hispanic_and_broad_race <- hispanic_and_broad_race_raw |>
    purrr::list_rbind() |>
    dplyr::inner_join(
        hispanic_and_broad_race_variables,
        by = "Index"
    ) |>
    dplyr::select(
        "Census Race",
        "Ethnicity",
        "Year",
        "Value"
    ) |>
    tidyr::pivot_wider(
        names_from = "Ethnicity",
        values_from = "Value",
        values_fill = 0
    ) |>
    dplyr::mutate(
        Total = .data$Hispanic + .data$`non-Hispanic`,
        `Percent Hispanic` = .data$Hispanic / .data$Total
    )
    
hispanic_and_broad_race |>
    dplyr::slice_sample(
        n = 1,
        by = "Year"
    ) |>
    dplyr::mutate(
        dplyr::across(c("Hispanic", "non-Hispanic", "Total"),
                      scales::label_comma(accuracy = 1)),
        dplyr::across("Percent Hispanic",
                      scales::label_percent(accuracy = 1))
    ) |>
    knitr::kable(
        align = "lrrrrr"
    )
```

Finally, we plot it

```{r}
#| label: plot-changes-in-hispanic-percentage
#| fig-dim: !expr "c(6, 6)"
#| dpi: 120

hispanic_and_broad_race |>
    dplyr::mutate(
        `Census Race` = stringr::str_to_title(.data$`Census Race`)
    ) |>
    ggplot2::ggplot(
        ggplot2::aes(x = .data$Year,
                     y = .data$`Percent Hispanic`,
                     color = .data$`Census Race`)
    ) +
    ggplot2::geom_line(
        linewidth = 1,
        lineend = "round",
        linejoin = "mitre"
    ) +
    ggplot2::geom_point(
        size = 3
    ) +
    ggplot2::scale_x_continuous(
        name = NULL,
        labels = scales::label_number(big.mark = ""),
        limits = c(2005, 2025),
        breaks = scales::breaks_width(5),
        minor_breaks = scales::breaks_width(1)
    ) +
    ggplot2::scale_y_continuous(
        name = "Percentage Hispanic",
        labels = scales::label_percent(accuracy = 1),
        limits = c(0, 1),
        breaks = scales::breaks_width(0.2),
        minor_breaks = scales::breaks_width(0.05)
    ) +
    ggplot2::scale_color_discrete(
        guide = ggplot2::guide_legend(
            title = NULL,
            position = "top",
            nrow = 4
        )
    ) +
    ggplot2::theme_minimal()
```

#### Methodological detail tables

Tables `B98013` and `B99021` give information about the Census's data collection
methods.
Statisticians can use to describe how much uncertainty there is in the data
concerning racial identities.
We can skip those.


## Appendix

### Subtables of race and ethnicity

```{r}
#| label: show-broad-subtables
#| echo: false
knitr::kable(broad_subtables)
```

### OMB Minimum Reporting Categories

| Minimum Race/Ethnicity Reporting Category | Definition |
|-------------------------------------------|------------|
| American Indian or Alaska Native          | Individuals with origins in any of the original peoples of North, Central, and South America, including, for example, Navajo Nation, Blackfeet Tribe of the Blackfeet Indian Reservation of Montana, Native Village of Barrow Inupiat Traditional Government, Nome Eskimo Community, Aztec, and Maya. |
| Asian | Individuals with origins in any of the original peoples of Central or East Asia, Southeast Asia, or South Asia, including, for example, Chinese, Asian Indian, Filipino, Vietnamese, Korean, and Japanese. |
| Black or African American | Individuals with origins in any of the Black racial groups of Africa, including, for example, African American, Jamaican, Haitian, Nigerian, Ethiopian, and Somali. |
| Hispanic or Latino | Includes individuals of Mexican, Puerto Rican, Salvadoran, Cuban, Dominican, Guatemalan, and other Central or South American or Spanish culture or origin. |
| Middle Eastern or North African | Individuals with origins in any of the original peoples of the Middle East or North Africa, including, for example, Lebanese, Iranian, Egyptian, Syrian, Iraqi, and Israeli. |
| Multiracial and/or Multiethnic | Those who identify with multiple race/ethnicity minimum reporting categories. |
| Native Hawaiian or Pacific Islander | Individuals with origins in any of the original peoples of Hawaii, Guam, Samoa, or other Pacific Islands, including, for example, Native Hawaiian, Samoan, Chamorro, Tongan, Fijian, and Marshallese. |
| White | Individuals with origins in any of the original peoples of Europe, including, for example, English, German, Irish, Italian, Polish, and Scottish. |

<!--->
## Separating race and ethnicity

This combination of categories is terrible for a variety of reasons.
Leaving aside the panoply of historical and ethical ones, there is also a purely
data one.
The table with not suffix letter holds the population for the whole area.
The different subtables will not necessarily add up to match that overall
population because there is overlap between the two groups.
Specifically, because these categories combine race and ethnicity, folks who are
counted in subtable "I" could also be counted in any subtable other than "H."

```{r}
#| label: compute-subgroupings
#| eval: false
#| echo: false

national_subgroupings <- national_populations |>
    dplyr::mutate(
        Nationwide = .data$Suffix == "",
        `Including only non-Hispanic whites (B:I)` = .data$Suffix %in% LETTERS[2:9],
        `Not including Hispanics (A:G)` = .data$Suffix %in% LETTERS[1:7]
    ) |>
    tidyr::pivot_longer(
        cols = c("Nationwide",
                 "Including only non-Hispanic whites (B:I)",
                 "Not including Hispanics (A:G)"),
        names_to = "Subgrouping",
        values_to = "Included"
    ) |>
    dplyr::filter(
        .data$Included
    ) |>
    dplyr::summarize(
        Population = sum(.data$Population),
        .by = "Subgrouping"
    )

national_subgroupings |>
    dplyr::arrange(
        dplyr::desc(.data$Population)
    ) |>
    dplyr::mutate(
        Population = scales::label_comma(accuracy = 1)(.data$Population)
    ) |>
    knitr::kable(
        align = "lr"
    )
```

Ok, so it looks like we can avoid double-counting, but only if we ignore
Hispanic ethnicity.

But, that means that we can count the total number of non-Hispanic, non-White
people.

```{r}
#| label: non-hispanic-non-white
#| eval: false
#| echo: false

hispanic_margins <- tibble::tribble(
    ~ Suffix, ~ Sign, ~ `Race/Ethnicity`,
    "I",           1, "Hispanic and not white",
    "A",          -1, "Hispanic and not white",
    "H",           1, "Hispanic and not white",
    "A",           1, "Hispanic and white",
    "H",          -1, "Hispanic and white",
    "I",           1, "Hispanic or Latino"
)

hispanic_populations <- national_populations |>
    dplyr::select(
        "Suffix",
        "Population",
    ) |>
    dplyr::inner_join(
        hispanic_margins,
        by = "Suffix"
    ) |>
    dplyr::summarize(
        Population = sum(.data$Population * .data$Sign),
        .by = "Race/Ethnicity"
    )

hispanic_populations |>
    dplyr::mutate(
        Population = scales::label_comma(accuracy = 1)(.data$Population)
    ) |>
    knitr::kable(
        align = "lr"
    )
```

I can't stop myself from pointing out that this table is only possible because
the Census decided that we just had to know which folks were not only white, but
also not Hispanic.

The ten categories that we see above are often the ten categories that the
Census uses to disaggregate their data.
As we saw above, that system does not give any information about how non-white
racial identities are distributed between Hispanic and non-Hispanic ethnic
identities.
There is another table that gives the breakdown in more detail, though:

```{r}
#| label: fetch-racine-race-counts-by-tract
#| eval: false
#| echo: false

racine_race_ethnicity_variables <- RACE_ETHNICITY_SUBTABLE_METADATA |>
    dplyr::filter(
        nchar(.data$Hispanic) > 0
    ) |>
    dplyr::select(
        "Census Race",
        tidyselect::ends_with("Hispanic")
    ) |>
    tidyr::pivot_longer(
        cols = tidyselect::everything(),
        names_to = "Ethnicity",
        values_to = "Variable"
    )

raw_racine_races <- racine_race_ethnicity_variables |>
    dplyr::pull(
        "Variable"
    ) |>
    fetch_data(
        year = 2022,
        for_geo = "tract",
        for_items = "*",
        survey_type = "acs",
        table_or_survey_code = "acs5",
        other_geos = list(state = 55L, 
                          county = 101L)
    )
```

```{r}
#| label: wrangle-racine-race-counts-by-tract
#| eval: false
#| echo: false
racine_races <- raw_racine_races |>
    dplyr::inner_join(
        ACS_RACE_ETHNICITY_VARIABLES,
        by = c("Group", "Index")
    )

racine_totals <- racine_races |>
    dplyr::summarize(
        Total = sum(.data$Value, na.rm = TRUE),
        .by = c("Ethnicity")
    )

knitr::kable(racine_totals)

county_subtotals <- racine_races |>
    dplyr::filter(
        .data$Race != "All"
    ) |>
    dplyr::summarize(
        Population = sum(.data$Value, na.rm = TRUE),
        .by = c("Race", "Ethnicity")
    ) |>
    dplyr::left_join(
        racine_totals,
        by = "Ethnicity"
    ) |>
    dplyr::mutate(
        Percentage = .data$Population / .data$Total
    )

knitr::kable(county_subtotals)
```
<!--->
