---
title: "Reusability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reusability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

commanism <- function(.) {
    scales::label_comma(accuracy = 1)(.)
}

```

## Introduction

A reusable approach incorporates setup and post-processing, which take place
before and after calling for data from the API.
Let's say that you are interested in the fraction of women who have given birth
in the U.S., broken down by age.

## The non-reusable part

Use `search_in_glossary()` to find tables that are relevant to our question.

```{r}
#| label: search-for-women-and-births

WB_TABLES <- hercacstables::search_in_glossary(
    hercacstables::METADATA_FOR_ACS_GROUPS,
    Group = "\\d$", # the group must end in a digit, so only "all races" tables
    Universe = "wom[ae]n", # matches woman and women
    Description = c("birth", "year") # the census doesn't use "age" much
)
```

```{r}
#| label: show-womens-and-birth-tables
#| echo: false
knitr::kable(WB_TABLES)
```

It looks like table
[B13016](https://api.census.gov/data/2023/acs/acs5/groups/B13016.html) is
exactly what we are looking for.

## Setup

### Define a glossary of variable meanings

First, create a glossary table that maps from the Census variables you need to
the real-world meanings that you actually care about.

The following example glosses 12 variables from "B13016."
Each specific variable encodes four columns' worth of data.

```{r}
#| label: setup-births-glossary
GLOSSARY_OF_WOMEN_AND_BIRTHS <- "B13016" |>
    hercacstables::unpack_group_details() |>
    dplyr::filter(
        .data$Dataset == "ACS1", # there are details for 1- and 5-year datasets
        dplyr::if_all(c("A", "B"), \(.)nchar(.) > 0) #
    ) |>
    dplyr::mutate(
        `Gave Birth` = stringr::str_detect(.data$A, "had"),
        `Lower Age` = stringr::str_extract(.data$B, "^\\d{2}"),
        `Upper Age` = stringr::str_extract(.data$B, "(?<!^)\\d{2}"),
        dplyr::across(tidyselect::ends_with("Age"),
                      as.integer)
    ) |>
    dplyr::select(
        !tidyselect::any_of(c("A", "B"))
    )
```

```{r}
#| label: show-births-glossary
#| echo: false

knitr::kable(GLOSSARY_OF_WOMEN_AND_BIRTHS)
```

## Calling the API

### Define a fetching function

Having defined a reusable glossary, we can now define a reusable fetching
function.

```{r}
#| label: define-fetch-births-data

fetch_women_and_births <- function(...) {
    hercacstables::fetch_data(
        variables = GLOSSARY_OF_WOMEN_AND_BIRTHS$Variable,
        survey_type = "acs",
        table_or_survey_code = "acs1",
        ...
    )
}
```

### Send the request to the API

Run the `fetch_data()` command by itself so that you can cache it.
In this example, we will ask for the most recent data for the whole country.

```{r}
#| label: fetch-births-data
#| cache: true

RAW_WOMEN_AND_BIRTHS <- fetch_women_and_births(
    year = hercacstables::most_recent_vintage("acs", "acs1"),
    for_geo = "us",
    for_items = "*"
)
```

```{r}
#| label: show-raw-birth-data
#| echo: false

RAW_WOMEN_AND_BIRTHS |>
    dplyr::mutate(
        Value = commanism(.data$Value)
    ) |>
    knitr::kable(
        align = "rlrrr"
    )
```

## Postprocessing

### Wrangle the result into a helpful format

The raw data do not answer our question all by themselves.
We are interested in rates, not counts.
Using the `dplyr::summarize()` function will let us get rid of the superfluous
information about marital status.

```{r}
#| label: wrangle-birth-data

wrangle_women_and_births <- function(.raw_api_output, ...) {
    .raw_api_output |>
        dplyr::inner_join(
            GLOSSARY_OF_WOMEN_AND_BIRTHS,
            by = c("Group", "Index")
        ) |>
        dplyr::summarize(
            `Recent Mothers` = sum(.data$Value * .data$`Gave Birth`),
            `All Women` = sum(.data$Value),
            Rate = .data$`Recent Mothers` / .data$`All Women`,
            .by = tidyselect::all_of(c(..., "Lower Age", "Upper Age"))
        )
}

WOMEN_AND_BIRTHS <- wrangle_women_and_births(RAW_WOMEN_AND_BIRTHS)
```

```{r}
#| label: show-wrangled-births
#| echo: false

WOMEN_AND_BIRTHS |>
    dplyr::mutate(
        dplyr::across(c("Recent Mothers", "All Women"),
                      commanism),
        Rate = scales::label_percent(accuracy = 1)(.data$Rate)
    ) |>
    knitr::kable(
        align = "r"
    )

```

## Reusability

The benefit of this approach is that we can reuse the fetching and wrangling
functions.

### Across Places

In this example, we pull data for three different counties in Wisconsin.

```{r}
#| label: pull-se-wi-births
#| cache: true

SE_WI_WOMEN_AND_BIRTHS <- fetch_women_and_births(
    state = 55,
    for_geo = "county",
    for_items = c("059", "079", "101"), # Kenosha, Racine, and Milwaukee
    year = 2023
) |>
    wrangle_women_and_births(
        "county"
    ) |>
    dplyr::mutate(
        county = dplyr::case_match(.data$county,
                                  "059" ~ "Kenosha",
                                  "079" ~ "Milwaukee",
                                  "101" ~ "Racine")
    )
```

For a simpler table, we'll exclude the counts of mothers and women, then pivot
the rate data by county.

```{r}
#| label: show-se-wi-wrangled-births
#| echo: false

SE_WI_WOMEN_AND_BIRTHS |>
    dplyr::select(
        "county",
        "Lower Age",
        "Upper Age",
        "Rate"
    ) |>
    dplyr::mutate(
        Rate = scales::label_percent(accuracy = 1)(.data$Rate)
    ) |>
    tidyr::pivot_wider(
        names_from = "county",
        values_from = "Rate"
    ) |>
    knitr::kable(
        align = "r"
    )
```

### Through time

In this example, we pull data about Texas from three years.
We have to make three separate calls, one for each year.
It's an idiosyncrasy of this particular API.

```{r}
#| label: pull-tx-births
#| cache: true

TEXAS_WOMEN_AND_BIRTHS <- 2021:2023 |>
    purrr::map(
        \(.y) fetch_women_and_births(
            for_geo = "state",
            for_items = 48L,
            year = .y
        )
    ) |>
    purrr::list_rbind() |>
    wrangle_women_and_births(
        "Year"
    )
```

For a simpler table, we'll exclude the counts of mothers and women, then pivot
the rate data by year.

```{r}
#| label: show-tx-wrangled-births
#| echo: false

TEXAS_WOMEN_AND_BIRTHS |>
    dplyr::select(
        "Year",
        "Lower Age",
        "Upper Age",
        "Rate"
    ) |>
    dplyr::mutate(
        Rate = scales::label_percent(accuracy = 1)(.data$Rate)
    ) |>
    tidyr::pivot_wider(
        names_from = "Year",
        values_from = "Rate"
    ) |>
    knitr::kable(
        align = "r"
    )
```
