---
title: "Children in Poverty"
---

```{r}
#| label: setup
#| include: FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  error = FALSE,
  message = FALSE
)
```

The two tracts are from the northern part of Nashua, NH.
Here's a map:

```{r}
#| label: map-of-nashville-blocks
#| fig-dim: !expr "c(8, 5)"
#| dpi: 72
#| echo: false

ggplot2::ggplot() +
    hercgeos::geom_terrain(
        dplyr::mutate(
            hercgeos::NASHVILLE_NH$blocks,
            Terrain = hercgeos::terrain_factor(.data$UR20,
                                               .water_area = .data$AWATER20,
                                               .land_area = .data$ALAND20)
        ),
        .line_color = "white"
    ) +
    ggplot2::geom_sf(
        data = hercgeos::NASHVILLE_NH$tracts,
        fill = NA,
        linewidth = 2
    ) +
    ggplot2::geom_sf_label(
        data = hercgeos::NASHVILLE_NH$tracts,
        mapping = ggplot2::aes(label = .data$GEOID)
    ) +
    ggspatial::annotation_scale(
        unit_category = "imperial",
        location = "br"
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(
        legend.position = "none"
    )
```

## Families by standard of living

The American Community Survey has a
[table](https://api.census.gov/data/2021/acs/acs5/groups/B17026.html)
that categorizes families according to the ratio between their income and the
[Federal poverty level in their community](https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-poverty-thresholds.html).
Nowadays, the Federal poverty level
[does not](https://www.latimes.com/opinion/story/2021-09-24/federal-poverty-level-us-families)
actually represent an income threshold where families really start struggling.
It is actually more of a measure of where you are in real danger of [starving to
death](https://www.ers.usda.gov/data-products/ag-and-food-statistics-charting-the-essentials/food-security-and-nutrition-assistance/).
For that reason, Higher Expectations for Racine County uses
**three times the Federal poverty level** as its threshold for a
[family-sustaining income](https://selfsufficiencystandard.org/wisconsin/).
The `hercacstables` package provides a table, `STANDARD_OF_LIVING_METADATA`,
that describes this classification.

```{r}
#| label: show-income-brackets
#| echo: false
hercacstables::STANDARD_OF_LIVING_METADATA |>
    dplyr::select(
        "Variable",
        "Least Poverty Ratio",
        "Greatest Poverty Ratio",
        "Standard of Living"
    ) |>
    knitr::kable(
        caption = "Income ratios from ACS table B17026, classified into standards of living"
    )
```

We use that table and the Census API look up how many families in each tract are
in each Census income bracket.

```{r}
#| label: lookup-standard-of-living
#| cache: true

RAW_STANDARD_OF_LIVING <- hercacstables::STANDARD_OF_LIVING_METADATA |>
    dplyr::pull(
        "Variable"
    ) |>
    hercacstables::fetch_data(
        year = 2021L,
        for_geo = "tract",
        for_items = c("010400", "010500"),
        survey_type = "acs",
        table_or_survey_code = "acs5",
        state = 33L,
        county = "011",
        use_key = TRUE
    )
```

Then, we aggregate to get counts of families experiencing "Unsustainable" and
"Sustainable" standards of living.

```{r}
#| label: wrangle-standard-of-living
#| echo: false
STANDARD_OF_LIVING <- RAW_STANDARD_OF_LIVING |>
    dplyr::inner_join(
        hercacstables::STANDARD_OF_LIVING_METADATA,
        by = "Index"
    ) |>
    dplyr::count(
        .data$tract,
        .data$`Standard of Living`,
        wt = .data$Value,
        name = "Families"
    ) |>
    tidyr::pivot_wider(
        names_from = "Standard of Living",
        values_from = "Families"
    ) |>
    dplyr::relocate(
        "tract",
        "Family-sustaining",
        "Unsustainable",
        "Everyone"
    ) |>
    dplyr::mutate(
        `Percent Family-sustaining` = .data$`Family-sustaining` / .data$Everyone
    )

STANDARD_OF_LIVING  |>
    dplyr::mutate(
        `Percent Family-sustaining` =
            scales::label_percent(accuracy = 1)(
                .data$`Percent Family-sustaining`
            )
    ) |>
    knitr::kable(
        align = "lrrrr"
    )
```

You can see that the economic situations in the two tracts are very different.

## Families by children present

We need to go from counts of families to counts of children, so we will next
look up how many families in each tract have children.
The census has a
[table](https://api.census.gov/data/2022/acs/acs1/groups/B11003.html) for this
that reports how many families have no kids, kids under 6, kids 6-17, or both.
The `hercacstables` package has a table, `FAMILIES_WITH_CHILDREN_METADATA`, that
describes the relationship between the ACS variables and the presence/absence of
children.

```{r}
#| label: show-families-with-children-metadata
#| echo: false
hercacstables::FAMILIES_WITH_CHILDREN_METADATA |>
    dplyr::select(
        "Variable",
        "Adults",
        "Children under 6",
        "Children 6-17"
    ) |>
    knitr::kable()
```

We can use this table to query the API to get the number of families with
children in each of the census tracts that we are interested in.
Notice that the only difference between this query and the previous one is which
variables we are asking for.
The whole point of the `hercacstables` package is to help you not repeat
yourself when you are asking for many different pieces of census information
about the same place.

```{r}
#| label: lookup-families-with-children
#| cache: true

RAW_FAMILIES_WITH_CHILDREN <- hercacstables::FAMILIES_WITH_CHILDREN_METADATA |>
    dplyr::pull(
        "Variable"
    ) |>
    hercacstables::fetch_data(
        year = 2021L,
        for_geo = "tract",
        for_items = c("010400", "010500"),
        survey_type = "acs",
        table_or_survey_code = "acs5",
        state = 33L,
        county = "011",
        use_key = TRUE
    )
```

We will aggregate the families into "Children Present" and "No Children."

```{r}
#| label: wrangle-families-with-children
#| echo: false

FAMILIES_WITH_CHILDREN <- RAW_FAMILIES_WITH_CHILDREN |>
    dplyr::inner_join(
        hercacstables::FAMILIES_WITH_CHILDREN_METADATA,
        by = "Index"
    ) |>
    dplyr::count(
        .data$tract,
        `Children` = dplyr::if_else(
            .data$`Children under 6` | .data$`Children 6-17`,
            "Children Present",
            "No Children"
        ),
        wt = .data$Value,
        name = "Families"
    ) |>
    tidyr::pivot_wider(
        names_from = "Children",
        values_from = "Families"
    ) |>
    dplyr::mutate(
        Total = .data$`Children Present` + .data$`No Children`,
        `Percent with Children` = .data$`Children Present` / .data$Total
    )

FAMILIES_WITH_CHILDREN |>
    dplyr::mutate(
        `Percent with Children` = scales::label_percent(accuracy = 1)(
            .data$`Percent with Children`
        )
    ) |>
    knitr::kable(
        align = "lrrrr"
    )
```

Once again, we see a pretty substantial difference between the two tracts.
Reassuringly, the total number of families is the same in both sets of data.

## Children per family by poverty level

The last set of raw data that we need is the number of children living in
poverty.
"But, wait!" you say, "Isn't that everything we needed in the first place?"
That would be true, except that the best we can get from the ACS is a 
[table](https://api.census.gov/data/2022/acs/acs1/groups/B05010.html)
that has detailed information about parents' birth origins but only distinguishes
three income tiers: less than the Federal poverty level, 1x-2x, and more than 2x
the Federal poverty level.
The `hercacstables` package's table `CHILDREN_IN_POVERTY_METADATA` describes how
the variables in that table map to parental place of birth and kid poverty.

```{r}
#| label: show-children-in-poverty-metadata
#| echo: false

hercacstables::CHILDREN_IN_POVERTY_METADATA |>
    dplyr::select(
        !c("Group", "Index")
    ) |>
    knitr::kable()
```

Here's the query to get those data:

```{r}
#| label: lookup-children-in_poverty
#| cache: true

RAW_CHILDREN_IN_POVERTY <- hercacstables::CHILDREN_IN_POVERTY_METADATA |>
    dplyr::pull(
        "Variable"
    ) |>
    hercacstables::fetch_data(
        year = 2021L,
        for_geo = "tract",
        for_items = c("010400", "010500"),
        survey_type = "acs",
        table_or_survey_code = "acs5",
        state = 33L,
        county = "011",
        use_key = TRUE
    )
```

We want **3x** the poverty level, so we will lump the "less than 1x" and "1x-2x"
tiers into "Unsustainable," leaving us with the "2x+" tier that mixes both the
"2x-3x" tier that we are calling "Unsustainable" and the "3x+" tier that we are
calling "Sustainable."

```{r}
#| label: wrangle-children-in-poverty
#| echo: false

CHILDREN_IN_POVERTY <- RAW_CHILDREN_IN_POVERTY |>
    dplyr::inner_join(
        hercacstables::CHILDREN_IN_POVERTY_METADATA,
        by = "Index"
    ) |>
    dplyr::count(
        .data$tract,
        .data$`Standard of Living`,
        wt = .data$Value,
        name = "Children"
    ) |>
    tidyr::pivot_wider(
        names_from = "Standard of Living",
        values_from = "Children"
    ) |>
    dplyr::mutate(
        Children = .data$Mixed + .data$Unsustainable
    )

CHILDREN_IN_POVERTY |>
    knitr::kable()
```

## Join the three datasets

Each of the three queries resulted in a data frame where rows correspond to
census tracts, so we can join the three results by the "tract" column.
We can also remove some of the duplicate columns to end up with just the tract,
the number of kids, the number of families, and the percentages of families with
children and with sustainable standards of living.

```{r}
#| label: join-the-data

CHILDREN_BY_TRACT <- CHILDREN_IN_POVERTY |>
    dplyr::inner_join(
        FAMILIES_WITH_CHILDREN,
        by = "tract"
    ) |>
    dplyr::inner_join(
        STANDARD_OF_LIVING,
        by = "tract"
    ) |>
    dplyr::select(
        "tract",
        "Children",
        "Mixed",
        "Observed Unsustainable" = "Unsustainable.x",
        "Families" = "Total",
        tidyselect::starts_with("Percent")
    )
```

This is what that table looks like:

```{r}
#| label: show-joined-tables
#| echo: false

CHILDREN_BY_TRACT |>
    dplyr::mutate(
        dplyr::across(tidyselect::starts_with("Percent"),
                      scales::label_percent(accuracy = 1))
    ) |>
    knitr::kable(
        align = "lrrrrrr"
    )
```

You can see that, in this case, we did not actually need to pull the number of
families with children.
If we were doing an even nerdier dive, though, we could use the fraction of
families with children to look at poverty across demographics like race or
ethnicity.

In this case, though, we just need to multiply the total number of children by
the percentage of families that have an sustainable standard of living.

```{r}
#| label: compute-sustainable-kids

SUSTAINABLE_KIDS <- CHILDREN_BY_TRACT |>
    dplyr::mutate(
        `Expected Unsustainable` = .data$Children * (1 - .data$`Percent Family-sustaining`),
        `Extra Unsustainable` = .data$`Expected Unsustainable` - .data$`Observed Unsustainable`,
        `Sustainable Kids` = .data$Mixed - .data$`Extra Unsustainable`,
        `Percent kids in Sustainable` = .data$`Sustainable Kids` / .data$Children,
        `Expected Sustainable` = .data$Children * .data$`Percent Family-sustaining`
    )

SUSTAINABLE_KIDS |>
    dplyr::select(
        "tract",
        "Children",
        "Percent Family-sustaining",
        "Expected Sustainable",
        "Expected Unsustainable",
        "Mixed",
        "Observed Unsustainable",
        "Extra Unsustainable",
        "Sustainable Kids",
        "Percent kids in Sustainable"
    ) |>
    knitr::kable()
```
