---
title: "Income Brackets for the Community Reinvestment Act"
author:
    - "Higher Expectations for Racine County"
    - "Ben Taft"
---

```{r}
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup

library(hercacstables)
```

## Context

The [Community Reinvestment Act](https://www.ffiec.gov/data/cra), or CRA, is
Federal legislation with mandates for how lenders must engage with folks living
in areas with relatively low incomes.
The Minneapolis Federal Reserve has really nice
[explanation](https://www.minneapolisfed.org/about-us/community-development-and-engagement/community-reinvestment-act) of it.
They even have a
[helpful article](https://www.minneapolisfed.org/article/2018/defining-low--and-moderate-income-and-assessment-areas)
about the definitions of "low and moderate income" with respect to a lender's
"assessment area."
These are key terms for place-based partnerships that want to demonstrate their
pertinence to a CRA lender.

The CRA defines four brackets of income level relative to an assessment area's
median income.
You can use `hercacstables` to pull an area's median income, define the bracket
ranges, and then to count the number of families in your area that fall into
each bracket.

## Areas

We need to identify two different geographical footprints for this analysis.
The first is the "assessment area," which is a broad geographical scope that
defines a baseline economic status for the CR.
The second is the specific area that your organization will impact.
This will be a subset of the assessment area, probably where poverty has been
concentrated by the systems that you are trying to change.

### Assesssment Area

Identifying the assessment area that is relevant to your organization is not
complicated: it is probably the Census-defined Metropolitan Statistical Area
where your organization is located.
If your organization serves part of an enormous metro area, you may need to
identify a specific Metropolitan Division.
If your organization serves a region with low population density, you may need
to use your entire state.
If you don't already know your area, you can start looking for it on this
[FAQ](https://www.census.gov/data/what-is-data-census-gov/guidance-for-data-users/frequently-asked-questions/how-do-i-search-by-address-using-data-census-gov.html).

For this article, we are going to use two Metropolitan Statistical Areas in
southeastern Wisconsin: Kenosha County and Racine County.

```{r}
#| label: assessment-areas

ASSESSMENT_AREAS <- tibble::tribble(
    ~ Area,    ~ county,
    "Kenosha", "059",
    "Racine",  "101"
)
```

### Impact Area

Your organization's impact area will be more customized to your case.
According to the article above, you should define it in terms of one or more
entire Census Tracts.

For this article, we are going to use two cities in southeastern Wisconsin:
Kenosha and Racine.

```{r}
#| label: impact-areas

IMPACT_AREAS <- tibble::tribble(
    ~ Area,    ~ city,  ~ district,
    "Kenosha", "39225", "07320",
    "Racine",  "66000", "12360"
)
```

## Incomes

The CRA wants us to look at income in three ways: the median in our assessment
area, the brackets defined by that median income, and the number of folks in
our assessment area in each bracket.

### Median income of assessment areas

We'll use table [`B19326`](https:://api.census.gov/data/acs/acs1/2024/groups/B19326.html),
which reports median income by sex and employment experience.
A full explanation of how I chose this table appears in an appendix at the
[end of this article](#Finding-a-Median-Income-Table).
We don't actually need any of the table's details, just the area's overall
median income.
That will be reported by the table's first row, `B19326_001E`.

```{r}
#| label: fetch-median-incomes

MEDIAN_INCOMES <- "B19326_001E" |>
    fetch_data(
        year = 2024L,
        for_geo = "county",
        for_items = ASSESSMENT_AREAS$county,
        survey_type = "acs",
        table_or_survey_code = "acs1",
        state = "55"
    ) |>
    dplyr::right_join(
        ASSESSMENT_AREAS,
        by = "county"
    ) |>
    dplyr::select(
        "Area",
        `Median Income` = "Value"
    )
```

```{r}
#| label: show-median-incomes
#| echo: false

MEDIAN_INCOMES |>
    dplyr::mutate(
        `Median Income` = scales::label_currency(accuracy = 1)(.data$`Median Income`)
    ) |>
    knitr::kable(
        caption = "Personal median incomes in 2024",
        align = "lr"
    )
```

### Defining income brackets

We can define a glossary of income brackets for each assessment area by joining
and multiplying.

```{r}
#| label: define-income-brackets

INCOME_BRACKETS <- tibble::tribble(
    ~ Bracket, ~ `Lower Bound`, ~ `Upper Bound`,
    "Low",                 0.0,             0.5,
    "Moderate",            0.5,             0.8,
    "Middle",              0.8,             1.2,
    "Upper",               1.2,             Inf
) |>
    dplyr::mutate(
        Bracket = forcats::fct_inorder(.data$Bracket)
    ) |>
    dplyr::cross_join(
        MEDIAN_INCOMES
    ) |>
    dplyr::mutate(
        dplyr::across(
            tidyselect::ends_with("Bound"),
            \(.) . * .data$`Median Income`
        )
    ) |>
    dplyr::select(
        "Area",
        "Bracket",
        "Lower Bound",
        "Upper Bound"
    )
```

```{r}
#| label: show-income-brackets
#| echo: false

INCOME_BRACKETS |>
    dplyr::mutate(
        dplyr::across(
            tidyselect::ends_with("Bound"),
            scales::label_currency(accuracy = 1)
        ),
        Range = paste(.data$`Lower Bound`,
                      .data$`Upper Bound`,
                      sep = " - ")
    ) |>
    dplyr::select(
        "Bracket",
        "Area",
        "Range"
    ) |>
    tidyr::pivot_wider(
        names_from = "Area",
        values_from = "Range"
    ) |>
    knitr::kable(
        caption = "Personal income brackets in 2024",
        align = "lrr"
    )
```

### Counting bracket populations

The third step is to count how many people fall into each income bracket.
To do this, we will use census table
[`B19325`](https://api.census.gov/data/2024/acs/acs1/groups/B19325.html)
`hercacstables` provides `GLOSSARY_OF_SEX_BY_INCOME`, which describes how each
row of this table corresponds to a specific combination of sex, employment
status, and income bracket.
We must assign each bracket in `B19325` to one or more of our local income
brackets, with the amount weighted by how much they overlap.

To do that, we'll create a helper function that computes how much, if at all,
two ranges overlap.
We'll use `purrr::map2_dbl` to make it work on vector inputs.

```{r}
#| label: define-overlap-helper

overlap_helper <- function(.lower_1, .upper_1, .lower_2, .upper_2) {
    .tops <- purrr::map2_dbl(.upper_1, .upper_2, min)
    .bottoms <- purrr::map2_dbl(.lower_1, .lower_2, max)
    purrr::map_dbl(.tops - .bottoms, \(.) max(., 0.0))
}
```

Now we join everything, compute the overlaps, filter out the rows without any
overlap, and compute the weights as the proportion of overlap.

```{r}
#| label: compute-brackets

GLOSSARY_OF_INCOME_BRACKETS <- GLOSSARY_OF_SEX_BY_INCOME |>
    dplyr::slice(
         -1L, # grand total
         -2L, # subtotal of all males
         -3L, # subtotal of full-time-employed males
         -5L, # subtotal of full-time-employed males with earnings
        -26L, # subtotal of not-full-time-employed males
        -28L, # subtotal of not-full-time-employed males with earnings
        -49L, # subtotal of all females
        -50L, # subtotal of full-time-employed females
        -52L, # subtotal of full-time-employed females with earnings
        -73L, # subtotal of not-full-time-employed females
        -75L  # subtotal of not-full-time-employed females
    ) |>
    dplyr::cross_join(
        INCOME_BRACKETS,
        suffix = c(" Census", " CRA")
    ) |>
    dplyr::mutate(
        Overlap = overlap_helper(.data$`Lower Bound Census`,
                                 .data$`Upper Bound Census`,
                                 .data$`Lower Bound CRA`,
                                 .data$`Upper Bound CRA`)
    ) |>
    dplyr::filter(
        .data$Overlap > 0.0
    ) |>
    dplyr::mutate(
        Proportion = dplyr::coalesce(
            .data$Overlap / 
                (.data$`Upper Bound Census` - .data$`Lower Bound Census`),
            1.0
        )
    ) |>
    dplyr::arrange(
        .data$Index,
        .data$Bracket,
        .data$Area
    )
```

```{r}
#| label: show-brackets
#| echo: false

GLOSSARY_OF_INCOME_BRACKETS |>
    dplyr::summarize(
        Indices = paste(c(min(.data$Index), max(.data$Index)),
                        collapse = " - "),
        .by = c("Area", "Bracket")
    ) |>
    tidyr::pivot_wider(
        names_from = "Bracket",
        values_from = "Indices",
        values_fill = "-"
    ) |>
    knitr::kable(
        align = "lrrrr"
    )
```

### Fetch the counts

The next step is to pull the counts of people in income brackets from the
Census.

```{r}
#| label: fetch-counts-by-bracket

RAW_COUNTS_BY_INCOME_BRACKET <- fetch_data(
    variables = "group(B19325)",
    year = 2024L,
    for_geo = "place",
    for_items = IMPACT_AREAS$city,
#    for_geo = "school district (unified)",
#    for_items = IMPACT_AREAS$district,
    survey_type = "acs",
    table_or_survey_code = "acs1",
    state = "55"
)
```

```{r}
#| label: wrangle-counts-by-bracket
COUNTS_BY_INCOME_BRACKET <- RAW_COUNTS_BY_INCOME_BRACKET |>
    dplyr::filter(
        .data$Measure == "E"
    ) |>
    dplyr::semi_join(
        GLOSSARY_OF_INCOME_BRACKETS,
        by = "Index"
    ) |>
    dplyr::inner_join(
        IMPACT_AREAS,
#        by = c(`school district (unified)` = "district")
        by = c(place = "city")
    ) |>
    dplyr::left_join(
        GLOSSARY_OF_INCOME_BRACKETS,
        by = c("Area", "Index"),
        relationship = "many-to-many"
    )
```

```{r}
#| label: compute-counts-by-bracket

POPULATIONS_PER_BRACKET <- COUNTS_BY_INCOME_BRACKET |>
    dplyr::summarize(
        People = sum(.data$Proportion * .data$Value, na.rm = TRUE),
        .by = c("Area", "Bracket")
    ) |>
    dplyr::arrange(
        .data$Area,
        .data$Bracket
    )
```

```{r}
#| label: show-count-by-bracket
#| echo: false

POPULATIONS_PER_BRACKET |>
    dplyr::mutate(
        Percent = .data$People / sum(.data$People),
        People = scales::label_comma(accuracy = 1)(.data$People),
        Percent = scales::label_percent(accuracy = 1)(.data$Percent),
        .by = "Area"
    ) |>
    knitr::kable(
        align = "llrr"
    )
```

The distributions of personal incomes differ between Kenosha and Racine.
The most populous income bracket in Kenosha, Upper, makes up 39% of people.
Kenosha`s next largest bracket, Lower income, also has a lot of people: 27%.
Its Moderate (14%) and Middle (20%) brackets are both less populous.
In contrast, Racine's Lower and Upper brackets are tied for most populous, at
29% each.
Racine's Middle (26%) bracket is also pretty populous, leaving the Moderate
income bracket as the smallest, with only 16% of the population.
Both income distributions are bimodal and asymmetric, with more folks at the
extreme ends of the income range.

## Appendix

If, like me, have NOT memorized every subtable of the ACS, you will need to
look up which tables report the data that you need.
The `search_in_columns` function and `METADATA_FOR_ACS_GROUPS` gloassary of
`hercacstables` exist for exactly this situation.

### Finding a Median Income Table

We definitely want to find tables with the word "median" in them.
The Census uses both "income" and "earnings" to talk about how much people make.
We'll search with both of those terms.

```{r}
#| label: look-up-median-income-tables

METADATA_FOR_ACS_GROUPS |>
    search_in_columns(
        Description = c("median", "income|earn"), # case-insensitive search for "median" AND ("income" OR "earn")
        Universe = "population",                  # this will give us individual incomes
        Group = "\\d$"                            # this will exclude tables that report within a single race
    ) |>
    knitr::kable()
```

Wow ... that's a lot of tables.
Poring through them, table
[`B19326`](https://api.census.gov/data/2024/acs/acs1/B19326.html)
seems promisingly broad.
In fact, if we cheat by looking [ahead](#Finding-counts-by-income) at tables
that count people by income, we find that there is a corresponding table
`B19325` that should make it easy to count people once we've determined
brackets.

Let's use `B19326` to calculate the
[median income of an assessment area](#Median-income-of-assessment-areas).

### Finding counts by income

Once again, we'll look for tables with either "income" or "earn" in their
descriptions.
This time, however, we want to specifically exclude tables that summarize
dollars.
We'll exclude tables with words like "median," "average," or "aggregate" in
their descriptions.

```{r}
#| label: look-for-income-counts

METADATA_FOR_ACS_GROUPS |>
    search_in_columns(
        Description = "income|earn",                          # seems obvious, right?
        `-Description` = c("aggregate", "median", "average"), # no summaries, we want counts of people
        Universe = "population",                              # this should give personal incomes
        Group = "\\d$"                                        # this will exclude tables that report within a single race
    ) |>
    knitr::kable()
```

We get a large number of tables, just like when we searched for
[tables of median incomes](#Finding-a-Median-Income-Table).
Once again, if we "cheat" by looking for tables that seem to go with each other,
three tables jump out:
[`B19325`](https://api.census.gov/data/2024/acs/acs1/B19325.html), 
[`B20001`](https://api.census.gov/data/2024/acs/acs1/B20001.html), and
[`B20005`](https://api.census.gov/data/2024/acs/acs1/B20005.html).
The difference between the three is that `B19325` counts people with any income,
`B20005` counts everyone over 16, and `B20001` only counts people with earnings.

Let's stick with income, rather than earnings.
We'll use `B19325` to [count](#Counting-bracket-populations) how many people
fall into each income bracket.

### Detailed bracket information

Here is the full table of each bracket from the Census and the proportion of
its members that go into each bracket from the CRA.

```{r}
#| label: view-counts-by-income-bracket
#| eval: true
COUNTS_BY_INCOME_BRACKET |>
    dplyr::select(
        "Area",
        "Sex",
        "Full-time",
        "Index",
        "Lower Bound Census",
        "Upper Bound Census",
        "Bracket",
        "Lower Bound CRA",
        "Upper Bound CRA",
        "Overlap",
        "Proportion"
    ) |>
    knitr::kable()
```

