---
title: "composition_of_income"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hercacstables)
```

The American Community Survey has twenty different tables that report on
aggregate household income.

```{r}
#| label: show-group-metadata
#| echo: false

METADATA_FOR_ACS_GROUPS |>
    dplyr::filter(
        .data$ACS5,
        stringr::str_starts(.data$Universe, "Household"),
        stringr::str_starts(.data$Description, "Aggregate")
    ) |>
    knitr::kable()
```

Only one of these tables, B19050, has more than one row.
It reports aggregate incomes by age of householder, with four age ranges that
break at 25, 45, and 64.
Its first rows is income over all ages, which is redundant with the single
row in table B19025.
Taken together, this gives us the 24 different variables that are described in
[hercacstables::GLOSSARY_OF_AGGREGATE_HOUSEHOLD_INCOME].

```{r}
#| label: show-glossary
#| echo: false

GLOSSARY_OF_AGGREGATE_HOUSEHOLD_INCOME |>
    knitr::kable()
```


