---
title: "Documenting Tenure, Age, and Years Living in Current Residence"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)

```

Some of the many things that
[The American Housing Survey](https://www.census.gov/programs-surveys/ahs.html)
reports about are
the age of the householder,
whether they rent or own their residence, and
how many years they have lived there.

This example considers how those factors interact.
It was motivated by interest in how often families with children at Racine
Unified School District have to move between schools.
Racine's decision-makers are particularly interested in understanding how
renting might impact student mobility.

## Setup

The data come from ACS table ["B25128"](https://api.census.gov/data/2024/acs/acs5/groups/B25128.html).

```{r}
#| label: show-glossary
#| echo: false

knitr::kable(hercacstables::GLOSSARY_OF_TENURE_DURATIONS_BY_AGE)
```

```{r}
#| label: define-number-of-big-units
#| echo: false

NUMBER_OF_BIG_UNITS <- 6
```

Here are the settings needed for pulling data for four different states.
For each state, we will look at the `r NUMBER_OF_BIG_UNITS` largest ones in the
state, focusing on one in particular from that subset.

```{r}
#| label: define-settings
#| echo: false

LOCATIONS <- tibble::tribble(
    ~ FIPS, ~ State,         ~ Focus,
       27L, "Minnesota",     "Minneapolis",
       33L, "New Hampshire", "Nashua",
       39L, "Ohio",          "Akron",
       55L, "Wisconsin",     "Racine"
)

QUERIES <- tibble::tribble(
    ~ Product, ~ Geography,                 ~ Label,
    "acs1",    "school district (unified)", "District",
    "acs1",    "place",                     "City"
)

SETTINGS <- dplyr::cross_join(LOCATIONS, QUERIES)

knitr::kable(SETTINGS)
```

```{r}
#| label: define-displacement-window
#| echo: false

DISPLACEMENT_WINDOW <- 2
```

If a household has moved in the last `r DISPLACEMENT_WINDOW` years then it
counts as a recent move.

Note that the AHS takes place on odd-numbered years.
That means that ACS data based on the AHS from even-numbered years are not
actually comparable with data from odd-numbered years.
Accordingly, this example only pulls ACS data from even-numbered years.
The current format of the AHS began in 2015, so this example only pulls data
from 2016 and later.

```{r}
#| label: define-acs-years
MOST_RECENT_ACS_YEAR <- hercacstables::most_recent_vintage("acs", "acs1")
ACS_YEARS <- seq(2016, MOST_RECENT_ACS_YEAR, 2)
```

## Fetch data

While the data always come from the same table, we will be querying the API
many times -- twice for each combination of survey year, geography, and state.
We could make fewer API calls per year if we downloaded data for EVERY state
each time, rather than a separate call for each state.
I have not profiled whether it is faster to do more small data pulls or fewer
lange data pulls.

There are two functions for fetching data.

### Function to Fetch Housing Data

Here is the function for pulling counts of households by
the age of the householder,
whether or not they rent, and
how recently they moved in.

```{r}
#| label: define-fetch-function
#| dependson:
#| - define-acs-years

fetch_tenure_durations <- function(Product, Geography, FIPS) {
    .years <- ACS_YEARS
    if (Product == "acs1") {
        .years <- .years[.years != 2020]
    }
    .years |>
        purrr::map(
            \(.year) hercacstables::fetch_data(
                hercacstables::GLOSSARY_OF_TENURE_DURATIONS_BY_AGE$Variable, 
                year = .year,
                survey_type = "acs",
                table_or_survey_code = Product,
                for_geo = Geography,
                for_items = "*",
                state = FIPS
            )
        ) |>
        purrr::list_rbind()
}
```

### Function to Fetch Populations and Names

Here is the function for pulling the names and populations of geographic units
in a state.
We will use these data to identify the largest units in each state.
Those are the units that we will use for comparison.

```{r}
#| label: define-fetch-names-and-sizes
#| dependson:
#| - define-acs-years

fetch_names_and_sizes <- function(Product, Geography, FIPS) {
    hercacstables::fetch_data(
        c("NAME", "B01001_001E"),
        year = MOST_RECENT_ACS_YEAR,
        for_geo = Geography,
        for_items = "*",
        survey_type = "acs",
        table_or_survey_code = Product,
        state = FIPS
    )
}
```

### Fetch the Data

This approach steps through each row of the SETTINGS table and fetches the
appropriate housing duration and population data.

```{r}
#| label: fetch
#| dependson:
#| - define-fetch-function
#| - define-fetch-names-and-sizes
#| - define-settings

RAW <- dplyr::mutate(
    SETTINGS,
    Durations = purrr::pmap(dplyr::pick("Product", "Geography", "FIPS"),
                            fetch_tenure_durations),
    Populations = purrr::pmap(dplyr::pick("Product", "Geography", "FIPS"),
                              fetch_names_and_sizes)
)

```

## Wrangle the Data

As always, there is a little aggregation to do with the data once we have pulled
what the ACS reports.
There are some computations to do for both types of data that we fetched.

### Define a Function to Wrangle Household Data

We will lump together the different durations of residency into two categories:
those who moved recently and
those who have not.

```{r}
#| label: define-wrangle-function
#| dependson:
#| - define-displacement-window

wrangle_tenure_duration <- function(Durations, Geography) {
    Durations |>
        dplyr::inner_join(
            hercacstables::GLOSSARY_OF_TENURE_DURATIONS_BY_AGE,
            by = c("Group", "Index")
        ) |>
        dplyr::summarize(
            Households = sum(.data$Value, na.rm = TRUE),
            `Recent Movers` = sum(
                .data$Value * (.data$`Longest Duration` < DISPLACEMENT_WINDOW),
                na.rm = TRUE
            ),
            `Recent Move Rate` = .data$`Recent Movers` / .data$Households,
            .by = tidyselect::all_of(c(
                "Year",
                Geography,
                "Tenure",
                "Upper Age"
            ))
        )
}

```

### Define a Function to Find the Largest Geographic Units

The full lists of geographic units from each state also need adjustments.
First, we simplify the names by removing redundant features like the state's
name and words like "school district."
Next, we identify which row corresponds to the focal geographic unit.
Finally, we pare down the number of columns to a useful minimum.

```{r}
#| label: define-wrangle-names-and-sizes
#| dependson:
#| - define-number-of-big-units

wrangle_names_and_sizes <- function(Populations, Geography, Focus, Label) {
    Populations |>
        dplyr::mutate(
            "{Label}" := .data$NAME |>
                stringr::str_extract(
                    "[^,]+"
                ) |>
                stringr::str_remove_all(
                    stringr::regex("city|metropolitan|municipal|district|public|schools?",
                                   ignore_case = TRUE)
                ) |>
                stringr::str_squish(),
            Focal = stringr::str_detect(.data[[Label]], Focus)
        ) |>
        dplyr::select(
            tidyselect::all_of(c(
                Label,
                Geography,
                "Focal",
                Population = "Value"
            ))
        ) |>
        dplyr::slice_max(
            .data$Population,
            n = NUMBER_OF_BIG_UNITS
        )
}
```

### Apply the Wrangling Functions

Once again, we apply the wrangling functions to each row of the raw data.
Once the duration and population data are wrangled, we join them into the
final data sets that we actually want to use.

```{r}
#| label: wrangle
#| dependson:
#| - fetch
#| - define-wrangle-function
#| - define-wrangle-names-and-sizes

WRANGLED <- dplyr::mutate(
    RAW,
    TenureDuration = purrr::pmap(
        dplyr::pick("Durations", "Geography"),
        wrangle_tenure_duration
    ),
    LargePlaces = purrr::pmap(
        dplyr::pick("Populations", "Geography", "Focus", "Label"),
        wrangle_names_and_sizes
    ),
    BigTenure = purrr::pmap(
        dplyr::pick("TenureDuration", "LargePlaces", "Geography"),
        \(TenureDuration, LargePlaces, Geography) {
            TenureDuration |>
                dplyr::inner_join(
                    LargePlaces,
                    by = Geography
                ) |>
                dplyr::arrange(
                    dplyr::across(tidyselect::all_of(Geography))
                )
        }
    )
)
```

## Summarize the Findings

Finally, we make some pretty tables and plots to display the results of all of
that wrangling.

### Define the Tabulation Function

This function makes a table where each cell shows both the percent recently
moved and the number of households in each combination of place and year.

```{r}
#| label: define-tabulate-function

tabulate_tenure_durations <- function(BigTenure, Label) {

    nyears <- length(unique(BigTenure$Year))
    
    BigTenure |>
        dplyr::summarize(
            `Recent Move Rate` = weighted.mean(
                .data$`Recent Move Rate`,
                .data$Households
            ),
            Households = sum(.data$Households),
            .by = tidyselect::all_of(c("Year", Label, "Tenure"))
        ) |>
        dplyr::mutate(
            dplyr::across("Recent Move Rate",
                          scales::label_percent(accuracy = 1)),
            dplyr::across("Households",
                          scales::label_comma(accuracy = 1)),
            Text = sprintf("%s of %s",
                           .data$`Recent Move Rate`,
                           .data$Households)
        ) |>
        dplyr::select(
            tidyselect::all_of(c("Year", Label, "Tenure", "Text"))
        ) |>
        tidyr::pivot_wider(
            names_from = "Year",
            names_sort = TRUE,
            values_from = "Text",
            values_fill = ""
        ) |>
        dplyr::arrange(
            dplyr::across(tidyselect::all_of(c(Label, "Tenure")))
        ) |>
        knitr::kable(
            caption = "Households that Have Recently Moved",
            align = paste0(c("l", rep("r", nyears)),
                           collapse = "")
        )
}
```

### Define the Plotting Function

This function makes line plots that show change in percentage recently moved
over time, with line colors by geographic unit and panels for each combination
of householder age and tenure.

```{r}
#| label: define-plot-function
#| dependson:
#| - define-displacement-window

plot_tenure_duration <- function(BigTenure, Label) {
    BigTenure |>
        ggplot2::ggplot(
            ggplot2::aes(x = .data$Year,
                         y = .data$`Recent Move Rate`,
                         color = .data[[Label]],
                         group = .data[[Label]],
                         alpha = .data$Focal)
        ) +
        ggplot2::geom_line(
            linewidth = 8.0 / 3.0,
            lineend = "round",
            linejoin = "round",
            na.rm = TRUE
        ) +
        ggplot2::scale_x_continuous(
            breaks = scales::breaks_width(2),
            minor_breaks = scales::breaks_width(1L),
            labels = scales::label_number(accuracy = 1,
                                          big.mark = "")
        ) +
        ggplot2::scale_y_continuous(
            name = paste("Moved in Last",
                         DISPLACEMENT_WINDOW,
                         "Years"),
            limits = c(0, 1),
            labels = scales::label_percent(accuracy = 1)
        ) +
        ggplot2::scale_alpha_manual(
            values = c(
                `TRUE` = 1,
                `FALSE` = 0.25
            ),
            guide = ggplot2::guide_none()
        ) +
        ggplot2::facet_grid(
            rows = ggplot2::vars(.data$`Upper Age`),
            cols = ggplot2::vars(.data$Tenure)
        ) +
        ggplot2::theme_minimal()
}
```

### Apply the Summary Functions

Another demonstration of how succinct code can be when using the functional
programming paradigm.

```{r}
#| label: summarize
#| dependson:
#| - wrangle
#| - define-tabulate-function
#| - define-plot-function

SUMMARIZED <- dplyr::mutate(
    WRANGLED,
    Table = purrr::pmap(dplyr::pick("BigTenure", "Label"),
                        tabulate_tenure_durations),
    Plot = purrr::pmap(dplyr::pick("BigTenure", "Label"),
                       plot_tenure_duration)
)
```

## Results

Here's what all of that prep creates:

```{r}
#| label: look-at-results
#| results: asis
#| dependson:
#| - summarize

SUMMARIZED |>
    tidyr::nest(
        .by = c("FIPS", "State", "Focus"),
        .key = "Results"
    ) |>
    purrr::pwalk(
        \(FIPS, State, Focus, Results) {
            cat(paste0("## ", Focus, ", ", State, "\n\n"))
            Results |>
                dplyr::select(
                    "Label", "Table", "Plot"
                ) |>
                purrr::pwalk(
                    \(Label, Table, Plot) {
                        cat("###", Label, "\n\n")
                        print(Table)
                        cat("\n\n")
                        print(Plot)
                        cat("\n\n")
                    }
                )
        }
    )
```
