---
title: "Plumbing and Owner-occupancy"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Let's say that you are a data intern for a [wastewater agency](https://cityofracine.org/WasteWater/Commission/) in the
southeastern part of Wisconsin.
Your mentors have asked you to find insights from the American Community Survey.

## Specifying your community

They tell you that the wastewater commission's area of responsibility is
essentially identical to the local school district's,
[Racine Unified](https://www.rusd.org).

You look for geographical levels related to schools in
`hercacstables::GLOSSARY_OF_ACS_GEOGRAPHIES`.

```{r}
#| label: look-for-school-geographies
SCHOOL_GEOGRAPHIES <- dplyr::filter(hercacstables::GLOSSARY_OF_ACS_GEOGRAPHIES,
                                    grepl("school", .data$`Geographic Level`))
```

This table tells us two things.
First, the geographic level that you should examine is "school district
(unified)".
Second, you also need to specify the state when you're fetching data about it.
With `hercacstables::fetch_data()`, and tools from the [tidyverse][tidyverse],
you can find the FIPS codes for unified school districts in Wisconsin that have
the word "Racine" in them.

```{r}
#| label: find-the-fips-for-racine-unified
RACINE_DISTRICTS <- "NAME" |>
    hercacstables::fetch_data(
        variables = _,
        year = 2020,                           # the most recent decennial
        for_geo = "school district (unified)", # see above
        for_items = "*",                       # this wildcard gets every item
        survey_type = "dec",                   # not actually the ACS *flex*
        table_or_survey_code = "pl",           # this data set is similar to ACS
        state = 55L                            # the Badger State
    ) |>
    dplyr::filter(
        stringr::str_detect(.data$NAME, "Racine")
    )
```

```{r}
#| echo: false

knitr::kable(RACINE_DISTRICTS)
```

## Finding relevant data

The first thing that you could do is to check for data related to plumbing.
If you were doing this without R, you might visit
[data.census.gov](https://data.census.gov) and search for "ACS plumbing".

### Groups with relevant data

With `hercacstables`, you would search within its glossary about ACS groups:

```{r}
#| label: search-for-group-about-plumbing

PLUMBING_GROUPS <- hercacstables::GLOSSARY_OF_ACS_GROUPS |>
    dplyr::filter(
        grepl("plumbing",         # search for the term "plumbing"
              .data$Description,  # within the "Description" column
              ignore.case = TRUE) # ignore capitalization
    )
```

```{r}
#| label: show-plumbing-groups
#| echo: false

knitr::kable(PLUMBING_GROUPS)
```

It appears that the keyword "plumbing" appears in `r nrow(PLUMBING_GROUPS)`
different ACS groups.

## Specific variables about plumbing

Let's say that "Tenure^[
Whether the occupants rent or own, not if they can't be fired.
] by plumbing facilities" jumps out to you as potentially interesting.

```{r}
#| label: define-group

GROUP <- "B25049"
```

There `hercacstables` also provides glossary about variables.
You can search it to see details about each variable in group `r GROUP`.

```{r}
#| label: search-for-variables-in-group

PLUMBING_VARIABLES <- hercacstables::GLOSSARY_OF_ACS_VARIABLES |>
    dplyr::filter(
        .data$Dataset == "ACS5", # use the 1-year dataset only
        .data$Group == GROUP     # pull all of the variables for this group
    )
```

```{r}
#| label: show-raw-plumbing-variables
#| echo: false

knitr::kable(PLUMBING_VARIABLES)
```

## Unpacking variable details

You can see that group `r GROUP` reports the number of households that have, or
do not have, plumbing facilities.
It further breaks those households down into renters and owner-occupants.
The raw glossary from the Census, and in `hercacstables::GLOSSARY_OF_ACS_VARIABLES`,
packs all of that information into the "Details" column.

It would be more useful if we actually had separate columns for the types of
tenure and plumbing facilities.
You can use `hercacstables::unpack_group_details()` to do this, in combination
with tools from the [tidyverse][tidyverse].

You might also decide that you don't the rows that report the total number of
households (row 1) and subtotals by tenure (2 and 5).
You'll have that data anyway from  3, 4, 6, and 7.

```{r}
#| label: make-a-helpful-glossary-table
PLUMBING_VARIABLES <- GROUP |>
    hercacstables::unpack_group_details() |>
    dplyr::filter(
        .data$Dataset == "ACS5"
    ) |>
    dplyr::rename(
        Tenure = "A",
        Plumbing = "B"
    ) |>
    dplyr::filter(
        dplyr::if_all(c("Tenure", "Plumbing"),
                      \(.) (nchar(.) > 0))
    )
```

```{r}
#| label: show-plumbing-variables
#| echo: false

knitr::kable(PLUMBING_VARIABLES)
```

## Fetch data

You are finally ready to pull a some census data about plumbing!
For starters, you could compare the local area to the whole county, the state,
its census region, and the whole country.
Once again, we'll use `hercacstables::fetch_data()` and the
[tidyverse][tidyverse].

```{r}
#| label: fetch-racine-plumbing
#| cache: true

YEAR <- hercacstables::most_recent_vintage("acs", "acs5")

RAW_PLUMBING <- tibble::tribble(
    ~ geo,                       ~ items, ~ other,
    "us",                        "1",     list(),
    "region",                    "2",     list(),
    "state",                     "55",    list(),
    "county",                    "101",   list(state = 55),
    "school district (unified)", "12360", list(state = 55)
) |>
    purrr::pmap(
        \(geo, items, other) rlang::exec(
            hercacstables::fetch_data,
            variables = c("NAME", PLUMBING_VARIABLES$Variable),
            year = YEAR,
            for_geo = geo,
            for_items = items,
            survey_type = "acs",
            table_or_survey_code = "acs5",
            !!!other
        )
    )
```

```{r}
#| label: show-raw-racine-plumbing
#| echo: false
#| results: asis

RAW_PLUMBING |>
    purrr::walk(
        \(.d) .d |>
            dplyr::slice(1) |>
            knitr::kable() |>
            print()
    )
```

The data in `RAW_PLUMBING` do not have much meaning without the information in
`PLUMBING_VARIABLES`.
Fortunately, we can use the [tidyverse][tidyverse] to join the two tables
together.

```{r}
#| label: wrangle-racine-plumbing

PLUMBING <- RAW_PLUMBING |>
    purrr::map(
        \(.) dplyr::select(.,
                           "Group",
                           "Index",
                           Geography = "NAME",
                           Households = "Value"
        )
    ) |>
    purrr::list_rbind() |>
    dplyr::right_join(
        PLUMBING_VARIABLES,
        by = c("Group", "Index")
    ) |>
    dplyr::select(
        "Geography",
        "Tenure",
        "Plumbing",
        "Households"
    )
```

```{r}
#| label: show_plumbing

knitr::kable(PLUMBING)
```

Finally, you can do some [tidyverse][tidyverse]-only magic to look at the
different rates of plumbing for renters versus owners.

```{r}
#| label: wrangle-rates-of-plumbing

RATES_OF_PLUMBING <- PLUMBING |>
    dplyr::mutate(
        dplyr::across(c("Tenure", "Plumbing"),
                      \(.) stringr::str_extract(., "^\\S+"))
    ) |>
    tidyr::pivot_wider(
        names_from = "Tenure",
        values_from = "Households"
    ) |>
    dplyr::mutate(
        `All Households` = .data$Owner + .data$Renter
    ) |>
    tidyr::pivot_longer(
        cols = c("Renter", "Owner", "All Households"),
        names_to = "Tenure",
        values_to = "Households"
    ) |>
    tidyr::pivot_wider(
        names_from = "Plumbing",
        values_from = "Households"
    ) |>
    dplyr::mutate(
        Households = .data$Complete + .data$Lacking,
        Rate = .data$Complete / .data$Households
    ) |>
    dplyr::select(
        "Geography",
        "Tenure",
        "Rate"
    ) |>
    tidyr::pivot_wider(
        names_from = "Tenure",
        values_from = "Rate"
    )
```

```{r}
#| label: show-rates-of-plumbing
#| echo: false

RATES_OF_PLUMBING |>
    dplyr::mutate(
        dplyr::across(!"Geography",
                      \(.) scales::label_percent(accuracy = 0.01)(.))
    ) |>
    knitr::kable(
        align = "lrrr"
    )
```

Three things might jump out at you.
The first is that almost every household in the US has complete plumbing
facilities.
The second thing is that plumbing access seems a little different for renters
and owners.
The third thing is that who has more access to plumbing is different in Racine
than at larger geographic levels.
At the scale of the nation, region, and state, renters are a little less likely
to have complete access to plumbing.
In contrast, renters in Racine County and the Racine Unified School District are
actually a little more likely to have access to plumbing.

[tidyverse]: https://www.tidyverse.org "The tidyverse"
